Completing correlation matrices

### 2021-11-22 paper/paper_risk.tex
Last small changes

\begin{abstract}
	We describe a way to complete a correlation matrix that is not fully specified. Such matrices arise in financial applications when the number of stochastic variables becomes large or when several smaller models are combined in a larger model. We argue that the proper completion to consider is the matrix that maximizes the entropy of the distribution described by the matrix. We then give a way to construct this matrix starting from the graph associated with the incomplete matrix. If this graph is chordal our construction will result in a proper correlation matrix. We demonstrate the construction using a cross-currency model with six stochastic variables.  
\end{abstract}

@@

\section{Introduction}
Modern financial models are driven by a large number of stochastic variables. These stochastic variables are usually correlated which creates the need to determine the coefficients of the underlying correlation matrix. Since the number of correlation coefficients increases quadratically with the number of stochastic variables, this number can become very large. While it would be desirable to calibrate all these coefficients this is in practice not possible because such a calibration procedure would be too slow, too unstable, or would require calibration instruments for which quotes can not be obtained. The issue of stability is especially important in a risk neutral setup. Daily calibrations might display undesired fluctuations if the calibration is not stable. Hybrid models are another source of undetermined correlation coefficients. Two calibrated models might be parts of a larger model. While the coefficients for the individual models have been determined in separate calibrations, the coefficients that mix variables from the two different models are undetermined. In a large model, with many smaller constituent models, the number of such coefficients can constitute the bulk of the matrix.

Faced with this situation, it is tempting to set the missing coefficients to zero. The idea here is that any other choice seems to add arbitrary information. This approach has several shortcomings though. The most important of which is that one might end up with a matrix that is not positive semidefinite and thus not a valid correlation matrix. To obtain a valid matrix further steps are required. One possibility is to use the fact that the set of positive definite correlation matrices is an open set that contains the unit matrix. It is thus always possible to scale the off-diagonal elements of the matrix with a common scale factor $\lambda<1$ and obtain a positive definite matrix. Another possibility is to choose the nearest correlation matrix in a suitable topology (see e.g \cite{higham}). These additional steps might change all the coefficients of the matrix, including those coefficients, that have been obtained in previous calibrations. They also rely on somewhat arbitrary further assumptions like a particular notion of distance on the space of matrices.

@@

We then show how to construct this completion. The construction starts by looking at the graph associated with the incomplete matrix. The vertices of the graph are the rows (or columns) of the matrix. Since we are looking at a correlation matrix, the vertices correspond to the stochastic variables of the model. Two vertices are connected by an edge in the graph if the corresponding correlation coefficient in the matrix is know. It is here that we encounter the one condition that needs to be satisfied for the construction to work. The graph that we have just constructed needs to be chordal. A graph is chordal if every loop in the graph that has more than three edges contains a chord, i.e. an edge that connects two non-adjacent vertices in the loop. If this condition is satisfied we can follow a construction first introduced by \cite{grone} and refined by \cite{smith} to complete the matrix. Every step of the construction adds new vertices to the edge which in turn correspond to entries in the correlation matrix. The matrix obtained in this way will automatically be a positive-semidefinite so that no further steps need to be taken to obtain a valid correlation matrix.

The paper now consists of three parts. In the next section we discuss the completion criterium. The matrix can be characterized in three different ways and we present them here. We then discuss the construction of the completion in section \ref{sec.procedure}. We do this by following the concrete example of a cross-currency model consisting of two interest rate models that are linked by one exchange rate. The two interest rate models and the model for the exchange rate are assumed to be stochastic volatility models so that we end up with a total of six stochastic variables. We will use this model to describe the all the required steps. These steps can easily be generalized to more complicated models. The $N$-currency model with $(N-1)$ foreign exchange rates is one such model. It is described in section \ref{sec.several}.

\section{The completion criterium}\label{sec.criterium}
Given a correlation matrix with unknown coefficients what is the right way to complete it? To motivate the answer we need to introduce some notation. Let 
\begin{equation}
	X = ( x_1, \ldots, x_k )^T\in \R^k
\end{equation}
be a $k$ dimensional random variable with a density function given by a $k$ dimensional normal distribution
\begin{align}
	f(X) & = N_k[\mu, H_X]( X ) \\
	& = (2\pi)^{-k/2}\det(H_X)^{-1/2}\exp\left(-\frac{1}{2}(X -\mu)^T H_X^{-1} (X-\mu)\right),
\end{align}
with mean $\mu$ and covariance matrix $H_X$. A crucial property of the normal distribution is that if we condition on some of the $x_i$, $i=1, \dots, k$, we again obtain a normal distribution. Let us assume that we want to condition on the last $l$ components of $X$:
\begin{equation}
	Z = (x_{k-l+1}, \ldots, x_k)^T \in \R^l.
\end{equation}
We partition the matrix $H$ in a way that reflects this partition of $X$:
\begin{equation}
	H_X = \begin{pmatrix}
		A & B \\
		B^T & C
	\end{pmatrix},
\end{equation}
with $A\in M_{k-l}$, $C\in M_l$, and $B\in M_{k-l,l}$. Let $\bar X$ be the first $k-l$ components of $X$:
\begin{equation}
	\bar X = (x_1, \ldots, x_{k-l})^T\in \R^{k-l}
\end{equation}
If we assume that $X$ has zero mean (i.e. $\mu=0$), the density function for $X$ conditioned on $Z$ is then given by
\begin{equation}
	f(\bar X \vert Z ) = N_{k-l}[BC^{-1}Z, H_X/C ](\bar X),
\end{equation}
(see \cite[chapter 8]{rao} and \cite{cottle}) where $H_X/C$ is the Schur complement of $C$ in $H_X$:
\begin{equation}\label{eqn.hxc}
	H_X/C = A - B C^{-1}B^T
\end{equation}
(Emilie Haynsworth introduced the name and highlighted its usefulness in \cite{hayns}. Schur originally made use of it in \cite{schur}. For an overview of the properties of the Schur complement see \cite{horn}.) We see that we again obtain a normal distribution. We will meet the Schur complement again soon.

@@

This choice of $W$ can be characterized in two more ways (see \cite{smith} and \cite{ksd}). Banachiewicz showed that the inverse of a matrix can be expressed using the Schur complement. Since the Schur complement $H/C$ is block diagonal the inverse of $H$ is given by
\begin{equation}
	H^{-1} = \begin{pmatrix}
		(H_X/C)^{-1} & -(H_X/C)^{-1} B C^{-1} & 0 \\
		- C^{-1} B^T (H_X/C)^{-1} & \Xi & - C^{-1} D (H_Y/C)^{-1} \\
		0 & - (H_Y/C)^{-1} D^T C^{-1} & (H_Y/C)^{-1} 
	\end{pmatrix},
\end{equation}
with
\begin{equation}
\Xi = C^{-1} + C^{-1} B^T (H_X/C)^{-1} B C^{-1} + C^{-1} D (H_Y/C)^{-1} D^T C^{-1}.
\end{equation}
We see that $H^{-1}$ has zeroes in the places where $W$ sits in $H$. The $W$ from equation (\ref{eqn.choice}) is the unique choice with that property.

@@

Our model will consist of two interest rate models; one for the domestic currency $E$ and one the foreign currency $A$. These two currencies are linked by an exchange rate $X$. We will model the two interest rates and the exchange rate with stochastic volatility models so that we have a total of six stochastic variables:
\begin{equation}\label{eqn.stochvar}
	E, \nu_E, A, \nu_A, X, \nu_X
\end{equation}
We will denote the corresponding correlation coefficients by 
\begin{equation}
	(a,b),
\end{equation}
where $a$ and $b$ are one of the stochastic variables from equation (\ref{eqn.stochvar}). In total there are 15 different correlation coefficients in the correlation matrix $H$ of our model. We assume that the coefficients 
\begin{equation}
	(E,\nu_E)\text{\ and\ }(A,\nu_A)
\end{equation}
have already been determined in separate calibrations of the interest rate models. That leaves us with 13 free coefficients in $H$. This number of coefficients is too large for a stable calibration. Instead we will only calibrate the coefficients
\begin{equation}
	(E,A), (E,X), (A,X),(X,\nu_X),
\end{equation}
and determine the remaining nine coefficients using the completion procedure described here (see figure \ref{fig.xccymodel}).

@@

\section{Conclusion}\label{sec.conclusion}
The increasing size of financial models has made insufficiently specified correlation matrices a common occurrence. The larger the model the fewer correlation coefficients are usually known. We thus need a rational for how to deal with these incomplete matrices as well as a robust method for implementing this rational. In this paper we have supplied both. We have argued that the correct way to complete the matrix is the one that maximizes the entropy of the distribution described by the matrix. This is the matrix with the maximal determinant. We have then shown how to actually construct this matrix with the maximal determinant. It is straight forward to implement the steps outlined in section \ref{sec.procedure}. Even for models with a large number of stochastic variables the procedure is both robust and fast.

@@

\begin{thebibliography}{MM}
	\bibitem[Higham, 2002]{higham}, Higham, N. J. (2002). Computing the nearest correlation matrix - a problem from finance. \textit{IMA Journal of Numerical Analysis} 22: 329 - 343.
	\bibitem[Grone et~al., 1984]{grone} Grone R., Johnson C.R., Sa E., Wolkowicz H. (1984). Positive definite completion of partial Hermitian matrices. \textit{Linear Algebra Appl.} 58: 109--124.
	\bibitem[Smith, 2008]{smith} Smith, R. L. (2008). The positive definite completion problem revisited. \textit{Linear Algebra and its Applications} 429:1442 -- 1452.	
    \bibitem[Georgescu, 2018]{georgescu} Georgescu, D. I., Higham, N. J. \& Peters, G. W. (2018). Explicit solutions to correlation matrix completion problems, with an application to risk management and insurance. \textit{Roy Soc Open Sci} 5: 172348.
	\bibitem[Rao, 2002]{rao} C. R. Rao (2002). \textit{Linear Statistical Inference and its Applications, 2nd edition}. John Wiley \& Sons, Inc.
	\bibitem[Cottle, 1974]{cottle} Cottle, R. W. (1974). Manifestations of the Schur complement. \textit{Linear Algebra and its Applications} 8:189--211. 
	\bibitem[Haynsworth, 1968]{hayns} Haynsworth, E.V. (1968), Determination of the inertia of a partitioned Hermitian matrix,  \textit{Linear  Algebra and its Applications} 1:73--81.
	\bibitem[Schur, 1917]{schur} Schur, J. (1917). \"Uber Potenzreihen, die im Innern des Einheitskreises beschr\"ankt sind. \textit{Journal f\"ur die reine und angewandte Mathematik} 147: 205--232.
	\bibitem[Horn, 2005]{horn} Horn, R. A. (2005). Basic  Properties  of  the  Schur  Complement, 17--46. In Fuzhen Zhang (Ed.) \textit{The Schur complement and its application}. Springer.
	% \bibitem[Dreyer et~al., 2021] {ksd} our other paper ...
	\bibitem[Golumbic, 2004]{golumbic} Golumbic, M. C. (2004). \textit{Algorithmic Graph Theory and Perfect Graphs}. Elsevier. 
	\bibitem[Blair et~al., 1993]{blair} Blair, J. R. S., Peyton, B. (1993). An introduction to chordal graphs and clique trees, 1 -- 29. In George, A., Gilbert, J. R., Liu, J. W. H. (Eds.). \textit{Graph Theory and Sparse Matrix Computation}, The IMA Volumes in Mathematics and its Applications 56. Springer.
\end{thebibliography}

>>>>

\begin{abstract}
	We describe a way to complete a correlation matrix that is not fully specified. Such matrices often arise in financial applications when the number of stochastic variables becomes large or when several smaller models are combined in a larger model. We argue that the proper completion to consider is the matrix that maximizes the entropy of the distribution described by the matrix. We then give a way to construct this matrix starting from the graph associated with the incomplete matrix. If this graph is chordal our construction will result in a proper correlation matrix. We give a detailed description of the construction for a cross-currency model with six stochastic variables and describe extensions to larger models involving more currencies.
\end{abstract}

@@

\section{Introduction}
Modern financial models are driven by a large number of stochastic variables. These stochastic variables are usually correlated which creates the need to determine the coefficients of the underlying correlation matrix. Since the number of correlation coefficients increases quadratically with the number of stochastic variables, this number can become very large. While it would be desirable to calibrate all these coefficients this is in practice not possible because such a calibration procedure would be too slow, too unstable, or would require calibration instruments for which quotes can not be obtained. The issue of stability is especially important in a risk neutral setup where daily calibrations might display undesired fluctuations if the calibration is not stable. Hybrid models are another source of undetermined correlation coefficients. Two calibrated models might be parts of a larger model. While the coefficients for the individual models have been determined in separate calibrations, the coefficients that mix variables from the two different models are undetermined. In a large model, with many smaller constituent models, the number of such coefficients can constitute the bulk of the matrix.

Faced with this situation, it is tempting to set the missing coefficients to zero. The idea here is that any other choice seems to add arbitrary information. This approach has several shortcomings though. The most important of which is that one might end up with a matrix that is not positive semidefinite and thus not a valid correlation matrix. To obtain a valid matrix, further steps are required. One possibility is to use the fact that the set of positive definite correlation matrices is an open set that contains the unit matrix. It is thus always possible to scale the off-diagonal elements of the matrix with a common scale factor $\lambda<1$ and obtain a positive definite matrix. Another possibility is to choose the nearest correlation matrix in a suitable topology (see e.g \cite{higham}). These additional steps might change all the coefficients of the matrix, including those coefficients, that have been obtained in previous calibrations. They also rely on somewhat arbitrary further assumptions like a particular notion of distance on the space of matrices.

@@

We then show how to construct this completion. The construction starts by looking at the graph associated with the incomplete matrix. The vertices of the graph are the rows (or columns) of the matrix. Since we are looking at a correlation matrix, the vertices correspond to the stochastic variables of the model. Two vertices are connected by an edge in the graph if the corresponding correlation coefficient in the matrix is known. It is here that we encounter the one condition that needs to be satisfied for the construction to work: The graph that we have just constructed needs to be chordal. A graph is chordal if every loop in the graph that has more than three edges contains a chord, i.e. an edge that connects two non-adjacent vertices in the loop. If this condition is satisfied we can follow a construction first introduced by \cite{grone} and refined by \cite{smith} to complete the matrix. Every step of the construction adds new vertices to the edge which in turn correspond to entries in the correlation matrix. The matrix obtained in this way will automatically be a positive-semidefinite so that no further steps need to be taken to obtain a valid correlation matrix.

The paper now consists of three parts. In the next section we discuss the completion criterium. The matrix can be characterized in three different ways and we present them here. We then discuss the construction of the completion in section \ref{sec.procedure}. We do this by following the concrete example of a cross-currency model consisting of two interest rate models that are linked by one exchange rate. The two interest rate models and the model for the exchange rate are assumed to be stochastic volatility models so that we end up with a total of six stochastic variables. We will use this model to describe all the required steps. These steps can easily be generalized to more complicated models. The $N$-currency model with $(N-1)$ foreign exchange rates is one such model. It is described in section \ref{sec.several}.

\section{The completion criterium}\label{sec.criterium}
Given a correlation matrix with unknown coefficients what is the right way to complete it? To motivate the answer we need to introduce some notation. Let 
\begin{equation}
	X = ( x_1, \ldots, x_k )^T\in \R^k
\end{equation}
be a $k$-dimensional random variable with a density function given by a $k$-dimensional normal distribution
\begin{align}
	f(X) & = N_k[\mu, H_X]( X ) \\
	& = (2\pi)^{-k/2}\det(H_X)^{-1/2}\exp\left(-\frac{1}{2}(X -\mu)^T H_X^{-1} (X-\mu)\right),
\end{align}
with mean $\mu$ and covariance matrix $H_X$. A crucial property of the normal distribution is that if we condition on some of the $x_i$, $i=1, \dots, k$, we again obtain a normal distribution. Let us assume that we want to condition on the last $l$ components of $X$:
\begin{equation}
	Z = (x_{k-l+1}, \ldots, x_k)^T \in \R^l.
\end{equation}
We partition the matrix $H$ in a way that reflects this partition of $X$:
\begin{equation}
	H_X = \begin{pmatrix}
		A & B \\
		B^T & C
	\end{pmatrix},
\end{equation}
with $A\in M_{k-l}$, $C\in M_l$, and $B\in M_{k-l,l}$. Let $\bar X$ be the first $k-l$ components of $X$:
\begin{equation}
	\bar X = (x_1, \ldots, x_{k-l})^T\in \R^{k-l}
\end{equation}
If we assume that $X$ has zero mean (i.e. $\mu=0$), the density function for $X$ conditioned on $Z$ is then given by
\begin{equation}
	f(\bar X \vert Z ) = N_{k-l}[BC^{-1}Z, H_X/C ](\bar X),
\end{equation}
(see \cite[chapter 8]{rao} and \cite{cottle}) where $H_X/C$ is the Schur complement of $C$ in $H_X$:
\begin{equation}\label{eqn.hxc}
	H_X/C = A - B C^{-1}B^T
\end{equation}
(Emilie Haynsworth introduced the name and highlighted its usefulness in \cite{hayns}. Schur originally made use of it in \cite{schur}. For an overview of the properties of the Schur complement see \cite{horn}.) We see that we again obtain a normal distribution. We will meet the Schur complement again soon.

@@

This choice of $W$ can be characterized in two more ways (see \cite{smith}). Banachiewicz showed that the inverse of a matrix can be expressed using the Schur complement (see \cite{horn}). Since the Schur complement $H/C$ is block diagonal the inverse of $H$ is given by
\begin{equation}
	H^{-1} = \begin{pmatrix}
		(H_X/C)^{-1} & -(H_X/C)^{-1} B C^{-1} & 0 \\
		- C^{-1} B^T (H_X/C)^{-1} & \Xi & - C^{-1} D (H_Y/C)^{-1} \\
		0 & - (H_Y/C)^{-1} D^T C^{-1} & (H_Y/C)^{-1} 
	\end{pmatrix},
\end{equation}
with
\begin{equation}
\Xi = C^{-1} + C^{-1} B^T (H_X/C)^{-1} B C^{-1} + C^{-1} D (H_Y/C)^{-1} D^T C^{-1}.
\end{equation}
We see that $H^{-1}$ has zeroes in the places where $W$ sits in $H$. The $W$ from equation (\ref{eqn.choice}) is the unique choice with that property.

@@

Our model will consist of two interest rate models; one for the domestic currency $E$ and one for the foreign currency $A$. These two currencies are linked by an exchange rate $X$. We will model the two interest rates and the exchange rate with stochastic volatility models so that we have a total of six stochastic variables:
\begin{equation}\label{eqn.stochvar}
	E, \nu_E, A, \nu_A, X, \nu_X
\end{equation}
We will denote the corresponding correlation coefficients by 
\begin{equation}
	(a,b),
\end{equation}
where $a$ and $b$ are one of the stochastic variables from equation (\ref{eqn.stochvar}). In total there are 15 different correlation coefficients in the correlation matrix $H$ of our model. We assume that the coefficients 
\begin{equation}
	(E,\nu_E)\text{\ and\ }(A,\nu_A)
\end{equation}
have already been determined in separate calibrations of the interest rate models. That leaves us with 13 free coefficients in $H$. This number of coefficients is too large for a stable calibration. Instead we will only calibrate the coefficients
\begin{equation}
	(E,A), (E,X), (A,X),(X,\nu_X),
\end{equation}
and determine the remaining nine coefficients using the completion procedure described here (see figure \ref{fig.xccymodel}).

@@

\section{Conclusion}\label{sec.conclusion}
The increasing size of financial models has made insufficiently specified correlation matrices a common occurrence. The larger the model the fewer correlation coefficients are usually known. We thus need a rational for how to deal with these incomplete matrices as well as a robust method for implementing this rational. In this paper we have supplied both. We have argued that the correct way to complete the matrix is to choose the matrix that maximizes the entropy of the distribution described by the matrix. This is the matrix with the maximal determinant. Then we have shown how to actually construct this matrix. The procedure that we have described in section \ref{sec.procedure} is straight forward to implement and is guaranteed to produce a valid correlation matrix provided one starts with a matrix whose associated graph is chordal. Even for models with a large number of stochastic variables the procedure is both robust and fast.

@@

\begin{thebibliography}{MM}
	\bibitem[Higham, 2002]{higham}, Higham, N. J. (2002). Computing the nearest correlation matrix - a problem from finance. \textit{IMA Journal of Numerical Analysis} 22: 329 - 343.
	\bibitem[Grone et~al., 1984]{grone} Grone R., Johnson C.R., Sa E., Wolkowicz H. (1984). Positive definite completion of partial Hermitian matrices. \textit{Linear Algebra Appl.} 58: 109--124.
	\bibitem[Smith, 2008]{smith} Smith, R. L. (2008). The positive definite completion problem revisited. \textit{Linear Algebra and its Applications} 429:1442 -- 1452.	
    \bibitem[Georgescu, 2018]{georgescu} Georgescu, D. I., Higham, N. J. \& Peters, G. W. (2018). Explicit solutions to correlation matrix completion problems, with an application to risk management and insurance. \textit{Roy Soc Open Sci} 5: 172348.
	\bibitem[Rao, 2002]{rao} C. R. Rao (2002). \textit{Linear Statistical Inference and its Applications, 2nd edition}. John Wiley \& Sons, Inc.
	\bibitem[Cottle, 1974]{cottle} Cottle, R. W. (1974). Manifestations of the Schur complement. \textit{Linear Algebra and its Applications} 8:189--211. 
	\bibitem[Haynsworth, 1968]{hayns} Haynsworth, E.V. (1968), Determination of the inertia of a partitioned Hermitian matrix,  \textit{Linear  Algebra and its Applications} 1:73--81.
	\bibitem[Schur, 1917]{schur} Schur, J. (1917). \"Uber Potenzreihen, die im Innern des Einheitskreises beschr\"ankt sind. \textit{Journal f\"ur die reine und angewandte Mathematik} 147: 205--232.
	\bibitem[Horn, 2005]{horn} Horn, R. A. (2005). Basic  Properties  of  the  Schur  Complement, 17--46. In Fuzhen Zhang (Ed.) \textit{The Schur complement and its application}. Springer.
	\bibitem[Golumbic, 2004]{golumbic} Golumbic, M. C. (2004). \textit{Algorithmic Graph Theory and Perfect Graphs}. Elsevier. 
	\bibitem[Blair et~al., 1993]{blair} Blair, J. R. S., Peyton, B. (1993). An introduction to chordal graphs and clique trees, 1 -- 29. In George, A., Gilbert, J. R., Liu, J. W. H. (Eds.). \textit{Graph Theory and Sparse Matrix Computation}, The IMA Volumes in Mathematics and its Applications 56. Springer.
\end{thebibliography}
